---
title: "Capstone 2 Data Only"
output: pdf_document
---

# Suicide and Smartphones Data Compilation 

This document is created simply to show how the dataset named "wide" was created that was used in the "Suicide and Smartphones" report. It does not have the sections and layout that are required in the second Harvard "Data Science" series capstone project 2. 

## Installing R packages needed

```{r message=FALSE, error=FALSE}
# installing the R packages needed
if(!require(readxl)) install.packages("readxl", repos = "http://cran.us.r-project.org")
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")

```

## Importing the data

```{r message=FALSE, error=FALSE}
# importing data
getwd()
Study_1970_2002 <- read_excel("Study 1970-2002 suicide rate by age group.xlsx")
n15_19 <- read_excel("CDC suicide rate ages 15-19 1999 to 2017.xlsx")
n10_14 <- read_excel("CDC Suicide rate ages 10-14 1999 to 2017.xlsx")
n10_24 <- read_excel("CDC suicide rate ages 10-24 1999 to 2017.xlsx")
n20_24 <- read_excel("CDC Suicide rate ages 20-24 1999 to 2017.xlsx")
Census_compint <- read_excel("computer and internet use 1984-2009.xlsx")
Pew_SM <- read_excel("Pew Social Media Use 2.xlsx")
all_suicides <- read_excel("CDC all age suicide rate 1999 to 2017.xlsx")
cell_own <- read_excel("Cell smart phone ownership.xlsx")

# importing stock market data

Dow <- read_excel("Stock Market Index Dow Data.xlsx")
SP <- read_excel("Stock Market Index SP 500 Data.xlsx")
nasdaq <- read_excel("Stock Market Index nasdaq data.xlsx")

```

## Stock Market Data Compilation

```{r message=FALSE, error=FALSE}
# examining the imported stock market data

head(Dow)
head(SP)
head(nasdaq)

# keeping only the necessary columns for the stock market data

Dow <- Dow[,c(1,6)]
nasdaq <- nasdaq[,c(1,6)]
SP <- SP[,c(1,6)]

# examining the new data frames

head(Dow)
head(SP)
head(nasdaq)

# creating a column for the year of the Date column

Dow <- Dow %>% mutate(Year = year(Date))
SP <- SP %>% mutate(Year = year(Date))
nasdaq <- nasdaq %>% mutate(Year = year(Date))

```
```{r message=FALSE, error= FALSE}

# examining the new market data frames

Dow
SP
nasdaq

# keeping only the year and adj close columns

Dow <- Dow[,c(2,3)]
SP <- SP[,c(2,3)]
nasdaq <- nasdaq[,c(2,3)]

Dow
SP
nasdaq
```

```{r message=FALSE, error= FALSE}

# making these data frames all numeric

SP$'Adj Close' <- as.numeric(SP$'Adj Close')
Dow$'Adj Close' <- as.numeric(Dow$'Adj Close')
nasdaq$'Adj Close' <- as.numeric(nasdaq$'Adj Close')

SP
Dow
nasdaq

```



```{r}
# grouping by year and summarizing the close amounts

Dow1 <- Dow %>% 
  group_by(Year) %>% 
  summarize(Avg_close = mean(`Adj Close`))
Dow1 

SP1 <- SP %>% 
  group_by(Year) %>% 
  summarize(Avg_close = mean(`Adj Close`))
SP1 

nasdaq1 <- nasdaq %>% 
  group_by(Year) %>% 
  summarize(Avg_close = mean(`Adj Close`))
nasdaq1 
```

```{r message = FALSE, error=FALSE}

# combining markets datasets

colnames(Dow1) <- c("Year", "Dow_close")
colnames(SP1) <- c("Year", "SP_Close")
colnames(nasdaq1) <- c("Year", "nasdaq_Close")

Dow1
SP1
nasdaq1

# joining the data frames to one named "markets"

markets <- inner_join(Dow1, SP1, by = "Year")
markets <- inner_join(markets, nasdaq1, by = "Year")
markets

```

```{r message = FALSE, error=FALSE}

# creating new column with mean of three market columns

markets1 <- markets %>% mutate(Mkt_close = (Dow_close + SP_Close + nasdaq_Close)/3)
markets1

# selecting just the year and average close in a new dataframe

markets <- markets1[,c(1,5)]
markets

# rounding the summary column

markets$Mkt_close <- round(markets$Mkt_close, 0)
markets

# renaming the second column to "Markets"

colnames(markets) <- c("Year", "Markets")
markets

```

## Combining other data

```{r message = FALSE, error=FALSE}

# combine 1999-2017 15-19 and 20-24 suicide rate data

wide <- full_join(n15_19, n20_24, by = "Year")
colnames(wide) <- c("Year", "15-19_rate", "20-24_rate")
wide

# add in 10-14 rate

wide <- full_join(wide, n10_14, by = "Year")
colnames(wide) <- c("Year", "15-19_rate", "20-24_rate", "10-14_rate")

# add in cell and smart phone ownership

wide <- full_join(wide, cell_own)
wide

# add in 1970-2002 study

wide <- full_join(wide, Study_1970_2002)
wide

# arrange wide by year

wide <- arrange(wide, Year)
wide

```

```{r message = FALSE, error=FALSE}

# adjusting Pew social media to be usable

Pew_SM
Pew_SM <- Pew_SM[,c(1,6)]

```

```{r message = FALSE, error=FALSE}
colnames(Pew_SM) <- c("Survey_Date", "Avg_SM_Use")
Pew_SM$Survey_Date <- as_date(Pew_SM$Survey_Date)
class(Pew_SM$Survey_Date)

Pew_SM

Pew_SM <- Pew_SM %>% mutate(Year = year(Survey_Date))
Pew_SM

Pew_SM1 <- Pew_SM[,c(2,3)]
Pew_SM1

# group by year and average rate

Pew_SM2 <- Pew_SM1 %>% group_by(Year) %>%
  summarize(Avg_SM_Use = mean(Avg_SM_Use))

Pew_SM2

# add social media data to the wide dataset 

wide <- full_join(wide, Pew_SM2)
wide



```

```{r message = FALSE, error=FALSE}

# add in market data

wide <- full_join(wide, markets)
wide

# add in computer and home internet use

colnames(Census_compint) <- c("Year", "Comp_athome", "Internet_athome")

wide <- full_join(wide, Census_compint)
wide
```

```{r message = FALSE, error=FALSE}

# creating column names

colnames(wide) <- c("Year", "15-19_rate", "20-24_rate", "10-14_rate", "Owns_Cellphone", "Owns_Smartphone", "15-24", "25-44", "45-64", ">65", "Avg_SM_Use", "Markets", "Comp_athome", "Internet_athome")



# export completed wide dataset

write.csv(wide, "wide.csv")


```



